---
title: "IDS investigation worksheet"
author: "KESC: Kirin Bhatnagar (s2778182), Elif Naz Kartal (s2791458), Shaun Gallacher (s2830596), & Collin Megson (s2829727)"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you *do not* need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}

# Needed for mapping
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Libraries
library(tidyverse)
library(tidymodels)
library(sf)
library(readxl)

```

```{r loading data}

# ca = council authority
# hbr = health board of residence
# EASR = standard rate for 100,000 people in the EU
activity_by_ca <- read.csv("data/hd_activitybyca.csv")
activity_by_hbr <- read.csv("data/hd_activitybyhbr.csv")
mortality_by_ca <- read.csv("data/hd_mortalitybyca.csv")
mortality_by_hbr <- read.csv("data/hd_mortalitybyhbr.csv")

# Source: https://data.spatialhub.scot/dataset/local_authority_boundaries-is/resource/d24c5735-0f1c-4819-a6bd-dbfeb93bd8e4
# Source for map must be cited in project
# Can also be found on a uk government website


# Retreiving scotland map with CAs

shapefile_path <- "data/Local_Authority_Boundaries_-_Scotland"
map <- st_read(shapefile_path)
mortality_ca_map <- st_read(shapefile_path)

```

Note that number of deaths and crude rate is na for any grouping of age other than "All"

```{r tidying}

#Selecting columns from mortality_by_ca
mortality_by_ca <- select(mortality_by_ca, Year, CA, AgeGroup, Sex, Diagnosis, NumberOfDeaths, CrudeRate, EASR)

#Filter out non-specific CAs
mortality_by_ca <- mortality_by_ca %>% filter(CA != "S92000003" & is.na(CA) == FALSE)

# Joining to complete map
mortality_ca_map <- left_join(mortality_by_ca, mortality_ca_map, by = c("CA" = "code"))
mortality_ca_map <- st_as_sf(mortality_ca_map, sf_column_name = "geometry")
# Note some areas have multiple data

#Creates new variable to be predicted
mortality_ca_map <- mortality_ca_map %>% mutate(
  deviance = CrudeRate - EASR
)

#Drops na for crude rate, should result in only 
mortality_ca_map <- mortality_ca_map %>% filter(is.na(CrudeRate)== FALSE)

#Looks at non-aged data and non-sexed data (inherent in EASR)
mortality_ca_map <- mortality_ca_map %>% filter(AgeGroup == "All", Sex == "All")



#Turns year into factor for dummy variable
mortality_ca_map <- mortality_ca_map %>% mutate(
  Year = as.factor(Year)
)

#Mutate into size types
map <- map %>% mutate(
  size_type = case_when(
    hectares<30000 ~ "small",
    hectares>30000 & hectares<90000 ~ "medium",
    hectares>90000 & hectares<300000 ~ "large",
    hectares>300000 ~ "very large"
  )
)

mortality_ca_map <- mortality_ca_map %>% mutate(
  size_type = case_when(
    hectares<30000 ~ "small",
    hectares>30000 & hectares<90000 ~ "medium",
    hectares>90000 & hectares<300000 ~ "large",
    hectares>300000 ~ "very large"
  )
)

#creates id
mortality_ca_map <- mortality_ca_map %>%
  mutate(id = row_number())
```

```{r example map visualizations}
# Use theme: theme(line = element_blank(),axis.text = element_blank(), panel.grid = element_line(color = "transparent"), panel.background = element_rect(fill='transparent'))

general_mortality_ca_exp <- mortality_ca_map %>% filter(Sex == "All", AgeGroup == "All", Year == "2018", Diagnosis == "Coronary Heart Disease") 

ggplot(general_mortality_ca_exp) + facet_grid(~Diagnosis)+ geom_sf(aes(fill = CrudeRate-EASR)) + scale_fill_gradient(low = "green",high = "red")+ theme(line = element_blank(),axis.text = element_blank(), panel.grid = element_line(color = "transparent"), panel.background = element_rect(fill='transparent'))

```

Possible hypothesis about rural vs urban areas given that dataset now has land size. Or simply predicting across diseases based on area. Maybe if difference from european average is constant per disease type for each council area.

```{r Example visualization}
ggplot(filter(mortality_ca_map, Year == 2018)) + facet_grid(~Diagnosis)+ geom_sf(aes(fill = CrudeRate-EASR)) + scale_fill_gradient2(low = "green", mid = "white", high = "red")+ theme(line = element_blank(),axis.text = element_blank(), panel.grid = element_line(color = "transparent"), panel.background = element_rect(fill='transparent'))

#CHD and Heart attack seem to have similar crude rate - EASR, while heart failure appears to be constant across all CAs, maybe we could categorize based on hectares into small medium and large council authorites and predict crude rate - EASR 
```

```{r Deviance visualization}

#Create visualization that tracks deviance per year
mortality_ca_map %>% filter(Sex == "All", AgeGroup == "All") %>% ggplot(mapping = aes(x= Diagnosis, y = deviance)) + facet_grid(~Year) + geom_boxplot()

#There seems to be a considerable trend in deviance over the years, so we will factor it into the recipe, appears to be the same spread in deviance over the years, maybe do difference between deviance and average deviance for the year

```


```{r workflow creation}


```

```{r}
#Error, does not have all geometry codes

```

```{r}

map %>% ggplot(mapping = aes(x = hectares, y=0)) + geom_point()

```

```{r hbr }
hbr_map <- st_read("data/SG_NHS_HealthBoards_2019.shp")

names(hbr_map)

mortality_hbr_map <- hbr_map %>%
  left_join(mortality_by_hbr, by = c("HBCode" = "HBR")) %>%
  st_as_sf(sf_column_name = "geometry")

geom_sf_text(aes(label = HBName), parse = FALSE)

general_mortality_hbr <- mortality_hbr_map %>% filter(Sex == "All", AgeGroup == "All", Year == "2018", Diagnosis == "Coronary Heart Disease")

ggplot(general_mortality_hbr) +
  geom_sf(aes(fill = CrudeRate)) +
  theme(
    line = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    panel.background = element_rect(fill = "transparent")
  )
```

```{r}
activity_by_hbr <- activity_by_hbr %>% 
  filter(AgeGroup != "All" & HBR != "S92000003") 
activity_by_hbr <- activity_by_hbr %>% 
  mutate(Diagnosis = factor(
    Diagnosis, 
    levels = c(
      "Coronary Heart Disease",
      "Heart Attack",
      "Heart Failure",
      "Angina")), 
    AgeGroup = factor(AgeGroup), 
    Sex = factor(Sex)) 

set.seed(1234) 
activity_by_hbr_split <- initial_split(activity_by_hbr) 
activity_by_hbr_train <- training(activity_by_hbr_split) 
activity_by_hbr_test <- testing(activity_by_hbr_split) 
diag_recipe <- recipe(Diagnosis ~ NumberOfDischarges + EASR, data = activity_by_hbr_train) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_zv(all_predictors()) %>% 
  step_normalize(all_numeric_predictors()) 

diag_model <- multinom_reg() %>% 
  set_engine("nnet") 
diag_wf <- workflow() %>% 
  add_recipe(diag_recipe) %>% 
  add_model(diag_model) 
diag_fit <- diag_wf %>% fit(activity_by_hbr_train)

glance(diag_fit)
```

```{r}
diag_test_preds <- predict(diag_fit, new_data = activity_by_hbr_test, type = "prob") %>% 
  bind_cols(activity_by_hbr_test) 

diag_test_preds %>% 
  roc_curve(truth = Diagnosis, starts_with(".pred_")) %>% 
  autoplot() 

diag_auc <- diag_test_preds %>% 
  roc_auc(truth = Diagnosis, starts_with(".pred_")) 
diag_auc
```

```{r}
# In this section, we will be exploring the question "To what extent can age and sex be effective in predicting the crude rate of deaths due to coronary heart disease across Scotland?"

# Filtering Data
chd <- mortality_by_hbr %>%
  filter(Diagnosis == "Coronary Heart Disease")

# Splitting to Training and Testing
set.seed(1234)
chd_split <- initial_split(chd)
chd_train <- training(chd_split)
chd_test <- testing(chd_split)

# Recipe
chd_rec <- recipe(CrudeRate ~ AgeGroup + Sex, data = chd_train) %>%
  step_dummy(all_nominal_predictors())

# Since we are trying to predict a continuous outcome, we have to use linear regression.

chd_mod <- linear_reg() %>%
  set_engine("lm")

chd_wf <- workflow() %>%
  add_recipe(chd_rec) %>%
  add_model(chd_mod)

chd_fit <- chd_wf %>% fit(data = chd_train)

chd_pred <- predict(chd_fit, chd_test) %>% bind_cols(chd_test)

metrics(chd_pred, truth = CrudeRate, estimate = .pred)

# What does this tell us? 

# We can see that our R^2 is very high, which means that our explanatory variables (Age Group and Sex) explain around 95% of the variation in our response variable (Crude Rate).

# The variation in the set of data is dominated by age group.

# RMSE and MAE are very large as well. On average, the predictions of the model are off by between 49 and 73 deaths per 100,000 of the population, which is huge but reasonable due to the fact that crude rates increase a lot by age. 
```

```{r}
# How much has the crude rate changed across age groups from 2010 to 2023? This is an intuitive question.

# Grouping by Year and Age Group

trend_over_years <- mortality_by_hbr %>%
  group_by(Year, AgeGroup) %>%
  summarise(
    mean_crude_rate = mean(CrudeRate, na.rm = TRUE),
    total_deaths = sum(NumberOfDeaths, na.rm = TRUE)
  ) %>%
  ungroup()

# Plotting the Graph

ggplot(trend_over_years,
       aes(x = Year,
           y = mean_crude_rate,
           color = AgeGroup)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Trends in Crude Rate by Age Group",
    x = "Year",
    y = "Mean Crude Rate",
    color = "Age Group"
  ) +
  theme_minimal()

# Calculating Percentage Changes

age_percentage_change <- trend_over_years %>%
  group_by(AgeGroup) %>%
  summarise(
    start = mean_crude_rate[Year == min(Year)],
    end = mean_crude_rate[Year == max(Year)],
    percentage_change = ((end - start) / start) * 100
  )

age_percentage_change
```

```{r}
# Does the crude rate change equally for men and women?

sex_trends <- mortality_by_hbr %>%
  group_by(Year, Sex) %>%
  summarise(
    mean_crude_rate = mean(CrudeRate, na.rm = TRUE),
  ) %>%
  ungroup()

lm_men <- lm(mean_crude_rate ~ Year, data = filter(sex_trends, Sex == "Males"))

lm_women <- lm(mean_crude_rate ~ Year, data = filter(sex_trends, Sex == "Females"))

coef(lm_men)
coef(lm_women)

# Women have a more negative slope, which means that there is a faster decline. This is important to note!

ggplot(sex_trends, aes(x = Year, y = mean_crude_rate, color = Sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Trends in Crude Rate by Sex",
    y = "Mean Crude Rate",
    x = "Year"
  ) +
  theme_minimal()

#looking at the crude rate ind vs looking at the deviance
#related responses.
#discuss pros and cons of discussing each of them
#highlight limitations of the model afterwards
#many of them ae categorical for the deviance
#crude rate other variables can be used

# But I want to see how it changes as women and men get older not compare years.
```

```{r}
age_sex_trends <- mortality_by_hbr %>%
  group_by(AgeGroup, Sex) %>%
  summarise(mean_crude_rate = mean(CrudeRate, na.rm = TRUE)) %>%
  ungroup()

age_sex_trends_filtered <- age_sex_trends %>%
  filter(AgeGroup != "All" & AgeGroup != "under75 years")

ggplot(age_sex_trends_filtered, aes(x = AgeGroup, y = mean_crude_rate, color = Sex, group = Sex)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Average Crude Rate by Age Group and Sex",
    x = "Age Group",
    y = "Crude Rate"
    ) +
  theme_minimal()

ggplot(age_sex_trends_filtered, aes(x = AgeGroup, y = mean_crude_rate, fill = Sex)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Crude Rate by Age Group and Sex",
    x = "Age Group",
    y = "Crude Rate",
    fill = "Sex"
    ) +
  theme_minimal()
```

```{r}

map %>% ggplot() + geom_sf(aes(fill = size_type)) + theme(line = element_blank(),axis.text = element_blank(), panel.grid = element_line(color = "transparent"), panel.background = element_rect(fill='transparent'))

#Splitting data
set.seed(2323)
mortality_ca_map_split <- initial_split(mortality_ca_map)
mortality_ca_map_train <- training(mortality_ca_map_split)
mortality_ca_map_test  <- testing(mortality_ca_map_split)

mortality_ca_rec1 <- recipe(deviance ~ hectares + Diagnosis + Year, data = mortality_ca_map_train) %>% step_dummy(Diagnosis, Year)

mortality_ca_rec2 <- recipe(deviance ~ size_type + Diagnosis + Year, data = mortality_ca_map_train) %>% step_dummy(Diagnosis, Year, size_type)

mortality_ca_rec3 <- recipe(deviance ~ CA + Diagnosis + Year, data = mortality_ca_map_train) %>% step_dummy(Diagnosis, Year, CA)

mortality_ca_mod <- linear_reg() %>% set_engine("lm")

mortality_ca_workflow1 <- workflow() %>% add_model(mortality_ca_mod) %>% add_recipe(mortality_ca_rec1)

mortality_ca_workflow2 <- workflow() %>% add_model(mortality_ca_mod) %>% add_recipe(mortality_ca_rec2)

mortality_ca_workflow3 <- workflow() %>% add_model(mortality_ca_mod) %>% add_recipe(mortality_ca_rec3)

ca_fit1 <- mortality_ca_workflow1 %>% fit(data = mortality_ca_map_train)
ca_fit2 <- mortality_ca_workflow2 %>% fit(data = mortality_ca_map_train)
ca_fit3 <- mortality_ca_workflow3 %>% fit(data = mortality_ca_map_train)

glance(ca_fit1)
glance(ca_fit2)
glance(ca_fit3)

#Creates prediction
mortality_ca_pred1 <- predict(ca_fit1, mortality_ca_map_test, type = "numeric") %>% bind_cols(select(mortality_ca_map_test, deviance, local_auth, Diagnosis, Year, id)) %>% mutate(model = "1")

mortality_ca_pred2 <- predict(ca_fit2, mortality_ca_map_test, type = "numeric") %>% bind_cols(select(mortality_ca_map_test, deviance, local_auth, Diagnosis, Year, id)) %>% mutate(model = "2")

mortality_ca_pred3 <- predict(ca_fit3, mortality_ca_map_test, type = "numeric") %>% bind_cols(select(mortality_ca_map_test, deviance, local_auth, Diagnosis, Year, id)) %>% mutate(model = "3")


#Binds model data together
mortality_ca_pred <- bind_rows(mortality_ca_pred1, mortality_ca_pred2, mortality_ca_pred3)

ggplot(data = mortality_ca_pred, mapping = aes(x = .pred, y = deviance-.pred)) + facet_grid(~model) + geom_point() + geom_hline(yintercept = 0)


#Binds id and joins
mortality_ca_pred_vis1 <- predict(ca_fit1, mortality_ca_map, type = "numeric") %>% bind_cols(select(mortality_ca_map, deviance, local_auth, Diagnosis, Year, geometry, id)) %>% mutate(model = "1")

mortality_ca_pred_vis2 <- predict(ca_fit2, mortality_ca_map, type = "numeric") %>% bind_cols(select(mortality_ca_map, deviance, local_auth, Diagnosis, Year, geometry, id)) %>% mutate(model = "2")

mortality_ca_pred_vis3 <- predict(ca_fit3, mortality_ca_map, type = "numeric") %>% bind_cols(select(mortality_ca_map, deviance, local_auth, Diagnosis, Year, geometry, id)) %>% mutate(model = "3")



mortality_ca_pred_vis <- bind_rows(mortality_ca_pred_vis1, mortality_ca_pred_vis2, mortality_ca_pred_vis3, (mortality_ca_map %>% select(deviance, local_auth, Diagnosis, Year, geometry, id)%>% mutate(.pred = deviance, model ="0"))) %>% st_as_sf(sf_column_name = "geometry")

mortality_ca_pred_vis %>% filter(Year == "2017") %>% ggplot() + facet_grid(Diagnosis ~ model)+  geom_sf(aes(fill = .pred)) + scale_fill_gradient2(low = "green", mid = "white", high = "red", na.value = "grey")+ theme(line = element_blank(),axis.text = element_blank(), panel.grid = element_line(color = "transparent"), panel.background = element_rect(fill='transparent'))
```


```{r more model evaluating}

ggplot(data = mortality_ca_pred, mapping = aes(x = .pred, y = deviance-.pred)) + facet_grid(Diagnosis~model) + geom_point() + geom_hline(yintercept = 0)
```

```{r}

library(tidymodels)
library(sf)
library(ggplot2)

mortality_hbr_map_train <- mortality_hbr_map_train %>% filter(AgeGroup == "All" & Sex == "All")

# Ensure EU_binary is a factor
mortality_hbr_map <- mortality_hbr_map %>% 
  mutate(EU_binary = ifelse(CrudeRate > EASR, 1, 0)) %>%
  mutate(EU_binary = factor(EU_binary))

# Split data (stratified to maintain positive/negative balance)
set.seed(2323)
mortality_hbr_map_split <- initial_split(mortality_hbr_map, strata = EU_binary)
mortality_hbr_map_train <- training(mortality_hbr_map_split)
mortality_hbr_map_test  <- testing(mortality_hbr_map_split)

# Recipes
mortality_hbr_rec1 <- recipe(EU_binary ~ hectares + Diagnosis + Year,
                             data = mortality_hbr_map_train) %>%
  step_dummy(all_nominal_predictors())


# Logistic regression model
mortality_hbr_mod <- logistic_reg() %>% set_engine("glm")

# Workflows
mortality_hbr_workflow1 <- workflow() %>% add_model(mortality_hbr_mod) %>% add_recipe(mortality_hbr_rec1)


# Fit models
hbr_fit1 <- mortality_hbr_workflow1 %>% fit(data = mortality_hbr_map_train)


# Predict probabilities
pred1 <- predict(hbr_fit1, mortality_hbr_map_test, type = "prob") %>% bind_cols(mortality_hbr_map_test)


# ROC curves (positive class = 1, above EU average)
roc1 <- roc_curve(pred1, truth = EU_binary, .pred_1, event_level = "second")


# Plot ROC curves individually
autoplot(roc1)


# Compute AUCs
mortality_auc1 <- pred1 %>% roc_auc(truth = EU_binary, .pred_1, event_level = "second")


# Print AUCs
mortality_auc1

```
```{r}
library(tidymodels)
library(sf)
library(ggplot2)

# Ensure EU_binary is a factor
mortality_ca_map <- mortality_ca_map %>% 
  mutate(EU_binary = ifelse(CrudeRate > EASR, 1, 0)) %>%
  mutate(EU_binary = factor(EU_binary))

# Split data (stratified to maintain positive/negative balance)
set.seed(2323)
mortality_ca_map_split <- initial_split(mortality_ca_map, strata = EU_binary)
mortality_ca_map_train <- training(mortality_ca_map_split)
mortality_ca_map_test  <- testing(mortality_ca_map_split)

# Recipes
mortality_ca_rec1 <- recipe(EU_binary ~ hectares + Diagnosis + Year,
                             data = mortality_ca_map_train) %>%
  step_dummy(all_nominal_predictors())

mortality_ca_rec2 <- recipe(EU_binary ~ size_type + Diagnosis + Year,
                             data = mortality_ca_map_train) %>%
  step_dummy(all_nominal_predictors())

mortality_ca_rec3 <- recipe(EU_binary ~ CA + Diagnosis + Year,
                             data = mortality_ca_map_train) %>%
  step_dummy(all_nominal_predictors())

# Logistic regression model
mortality_ca_mod <- logistic_reg() %>% set_engine("glm")

# Workflows
mortality_ca_workflow1 <- workflow() %>% add_model(mortality_ca_mod) %>% add_recipe(mortality_ca_rec1)
mortality_ca_workflow2 <- workflow() %>% add_model(mortality_ca_mod) %>% add_recipe(mortality_ca_rec2)
mortality_ca_workflow3 <- workflow() %>% add_model(mortality_ca_mod) %>% add_recipe(mortality_ca_rec3)

# Fit models
ca_fit1 <- mortality_ca_workflow1 %>% fit(data = mortality_ca_map_train)
ca_fit2 <- mortality_ca_workflow2 %>% fit(data = mortality_ca_map_train)
ca_fit3 <- mortality_ca_workflow3 %>% fit(data = mortality_ca_map_train)

# Predict probabilities
pred1 <- predict(ca_fit1, mortality_ca_map_test, type = "prob") %>% bind_cols(mortality_ca_map_test)
pred2 <- predict(ca_fit2, mortality_ca_map_test, type = "prob") %>% bind_cols(mortality_ca_map_test)
pred3 <- predict(ca_fit3, mortality_ca_map_test, type = "prob") %>% bind_cols(mortality_ca_map_test)

# ROC curves (positive class = 1, above EU average)
roc1 <- roc_curve(pred1, truth = EU_binary, .pred_1, event_level = "second")
roc2 <- roc_curve(pred2, truth = EU_binary, .pred_1, event_level = "second")
roc3 <- roc_curve(pred3, truth = EU_binary, .pred_1, event_level = "second")

# Plot ROC curves individually
autoplot(roc1)
autoplot(roc2)
autoplot(roc3)

# Compute AUCs
mortality_auc1 <- pred1 %>% roc_auc(truth = EU_binary, .pred_1, event_level = "second")
mortality_auc2 <- pred2 %>% roc_auc(truth = EU_binary, .pred_1, event_level = "second")
mortality_auc3 <- pred3 %>% roc_auc(truth = EU_binary, .pred_1, event_level = "second")

# Print AUCs
mortality_auc1
mortality_auc2
mortality_auc3
```

